---
title: "The families and their parameters"
description: >
  How the two families are constructed, what each distributional parameter
  means, how to constrain the two excess dispersions to be equal, and a fitted
  demonstration that they can be told apart when they are not.
bibliography: ../references.bib
link-citations: true
---

```{r}
#| label: setup
#| include: false
# This article is PRECOMPILED (see vignettes/precompile.R) because it fits a
# model. Figures are generated once, here, and committed to
# vignettes/articles/figure/ alongside the .Rmd -- pkgdown does not re-run the
# chunks, so the images must already exist.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  dev      = "svg",
  fig.path = "figure/"
)
has_stan <- requireNamespace("cmdstanr", quietly = TRUE) ||
  requireNamespace("rstan", quietly = TRUE)
backend <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"
```

```{r}
#| label: libraries
#| message: false
#| warning: false
library(brms)
library(bicountbrms)
```

This article describes the parameters that are fitted. The companion article
[The anatomy of a paired count](paired-count-anatomy.html) describes the
coordinates in which they are interpreted, and
[Choosing priors](choosing-priors.html) gives the recipes for both.

## Trivariate reduction

Both families are constructed by trivariate reduction: three independent latent
counts are drawn, and one of the three enters both observed counts. @holgate1964
and @karlis2003 use the construction with Poisson components;
@kirkpatrickApplyingMultivariateDiscrete2016 and
@kirkpatrickRMKdiscreteSundryDiscrete2022 with negative-binomial components.

$$
y_1 = N_{\text{shared}} + N_1
\qquad
y_2 = N_{\text{shared}} + N_2
$$

`N_shared` is never observed. The likelihood sums it out analytically over
$k = 0, \ldots, \min(y_1, y_2)$,

$$
P(y_1 = x,\ y_2 = y) = \sum_{k=0}^{\min(x,y)}
  f_{\text{s}}(k)\, f_1(x-k)\, f_2(y-k),
$$

where $f_{\text{s}}$, $f_1$ and $f_2$ are the probability mass
functions of the three latent counts. With Poisson components, this sum is
Campbell's [-@campbell1934] equation (2.2). The first source is the modelled
response; the second is supplied as supplementary integer data through
`vint()`, because `brms::custom_family()` declares a single response column.

The shared latent count enters both observed counts and therefore induces their
covariance, $\mathrm{Cov}(y_1, y_2) = \mathrm{Var}(N_{\text{shared}})$. It
cancels from the difference,

$$
d = y_1 - y_2 = N_1 - N_2,
$$

so the difference depends on the two source-specific counts alone. For Poisson
latent counts, that difference is exactly Skellam-distributed [@skellam1946],
which is why `bipois()` induces the Skellam as its difference model.

brms requires one distributional parameter of every family to be named `mu`. In
these families, `mu` is the rate of the shared latent count. Since
$\mathrm{E}[y_1] = \mu + \lambda_1$, `mu` is not the mean of either observed
count, nor of their difference. The constraint is `custom_family()`'s rather
than a modelling choice, and the same applies to the spelling of the two
source-specific rates: `custom_family()` rejects a distributional parameter
name ending in a digit, so the code writes `lambdaone` and `lambdatwo` where the
documentation writes $\lambda_1$ and $\lambda_2$.

## Poisson or negative-binomial latent counts

`bipois()` draws each latent count from a Poisson distribution, so the
variance is equal to the mean. `binegbin()` draws each from NB2 instead and
estimates three scalar dispersions. NB2$(m, \phi)$ is Stan's
`neg_binomial_2` and R's `dnbinom(size = phi, mu = m)`: it has mean $m$ and
variance $m + m^2/\phi$, so a larger $\phi$ means less overdispersion,
and the Poisson is the $\phi \to \infty$ limit.

| dpar | latent count | rate | dispersion |
|---|---|---|---|
| `mu`, `shapes` | $N_{\text{shared}}$ | $\mu$ | $\phi_{\text{s}}$ |
| `lambdaone`, `shapexone` | $N_1$ | $\lambda_1$ | $\phi_{x1}$ |
| `lambdatwo`, `shapextwo` | $N_2$ | $\lambda_2$ | $\phi_{x2}$ |

`bipois()` takes the three rates alone; `binegbin()` takes all six. Every
parameter uses a log link.

The moments follow directly from the decomposition:

$$
\mathrm{Var}(y_1) =
  \Big(\mu + \tfrac{\mu^2}{\phi_{\text{s}}}\Big) +
  \Big(\lambda_1 + \tfrac{\lambda_1^2}{\phi_{x1}}\Big),
\qquad
\mathrm{Var}(d) =
  \Big(\lambda_1 + \tfrac{\lambda_1^2}{\phi_{x1}}\Big) +
  \Big(\lambda_2 + \tfrac{\lambda_2^2}{\phi_{x2}}\Big).
$$

The covariance involves the shared latent count alone, so neither excess
dispersion appears in it:
$\mathrm{Cov}(y_1, y_2) = \mu + \mu^2/\phi_{\text{s}}$
[@kirkpatrickApplyingMultivariateDiscrete2016]. Freeing the two
excess dispersions therefore changes both margins and the difference, and
leaves the covariance unchanged; the correlation moves only through its
denominator.

Two features of that covariance bound what the construction can represent. The
covariance is a variance, so it cannot be negative: both families represent
non-negative association only. It is also bounded above, because the shared
count contributes to both observed variances,

$$
0 \le \rho \le
  \min\!\left(\sqrt{V_1 / V_2},\ \sqrt{V_2 / V_1}\right),
$$

for $V_1$ and $V_2$ the variances of the two observed counts, with the
upper bound attained only when one source has no private component. In the
Poisson case, each variance equals its mean, and this is the ceiling Holgate
[-@holgate1964] states and calls the drawback. A construction admitting negative
correlation gives up the shared latent count, which is the quantity these
families estimate.

An observation-level random effect on the source-specific components would be
the obvious alternative to a scalar dispersion, and it fails synthetic
recovery. With one observed pair per unit but three latent deviates per unit,
the excess deviates absorb the residual: in the motivating dataset the
population standard deviation of the excess was recovered as 0.37 against a
true 0.85, and drawing fresh deviates gave $\mathrm{Var}(d) = 2.9$ against a
true 19.2. A conditional posterior-predictive check does not expose that
failure; a marginal one, drawing fresh deviates, does.

## Tying the two excess dispersions

Both negative-binomial constructors estimate a separate dispersion for each
source-specific component. The two sources are different instruments, and no
argument requires their source-specific excess to be equally overdispersed. Up
to release 0.9.1, a single distributional parameter `shapex` imposed
$\phi_{x1} = \phi_{x2}$; 0.10.0 frees the constraint for both constructors.

To impose that constraint deliberately, route both distributional parameters
through one non-linear parameter:

```{r}
#| label: tied-formula
#| eval: false
bf(y1 | vint(y2) ~ 1, mu ~ 1,
   nlf(lambdaone ~ lamx), nlf(lambdatwo ~ lamx),
   nlf(shapexone ~ shapexx), nlf(shapextwo ~ shapexx),
   lamx ~ 1, shapes ~ 1, shapexx ~ 1, nl = TRUE)
```

The generated Stan shows the tie directly. Both distributional parameters are
assigned from the same non-linear predictor, so the model contains one free
dispersion rather than two:

```
shapexone[n] = exp(nlp_shapexx[n]);
shapextwo[n] = exp(nlp_shapexx[n]);
```

The tie is worth checking in the generated code rather than assuming, because an
untied model would still compile and still sample: it would quietly fit the
six-parameter model the formula was meant to constrain.
`tests/testthat/test-stancode-shape.R` pins both lines.

A prior on the tied parameter is written with `class = "b"` and
`nlpar = "shapexx"`. Both fields differ from the pre-0.10.0 spelling, which was
`class = "Intercept"` with `dpar = "shapex"`. A prior written the old way names
no parameter in the tied model and is dropped without a warning, which leaves
the dispersion improper; see [Choosing priors](choosing-priors.html).

Under full pairing, every row informs both dispersions. Under partial
observation, the two are not equally identified: `shapextwo` governs the
always-observed margin and appears on both branches of the likelihood, while
`shapexone` appears on the matched branch alone. [A worked partially observed
fit](partially-observed-fit.html) shows what follows from that asymmetry.

## Discrimination of two dispersions an order of magnitude apart

The get-started vignette recovers all six distributional parameters from a
fully paired design. The question here is narrower: when $\phi_{x1}$ and
$\phi_{x2}$ differ substantially, does the fit resolve them as different, or
merely bracket each of them widely enough to contain the truth?

The simulation below sets the two an order of magnitude apart, at 0.7 and 6.0,
and gives the design 300 rows across ten vessels. No observation flag appears
anywhere: this is `binegbin()` with `vint(y2)` alone.

```{r}
#| label: disc-sim
set.seed(20260825)

n_vessel     <- 10L
n_per_vessel <- 30L
n            <- n_vessel * n_per_vessel

truth <- list(
  log_mu_int = log(6),
  sd_vessel  = 0.3,
  lone       = 6,
  ltwo       = 6,
  shapes     = 3,
  shapexone  = 0.7,   # first source: strongly overdispersed
  shapextwo  = 6.0    # second source: mildly overdispersed
)

vessel <- rep(seq_len(n_vessel), each = n_per_vessel)
mu_i   <- exp(truth$log_mu_int + truth$sd_vessel * rnorm(n_vessel)[vessel])

# One shared count per row, drawn once and entering both observed counts.
n_shared <- rnbinom(n, size = truth$shapes, mu = mu_i)

dat <- data.frame(
  y1     = n_shared + rnbinom(n, size = truth$shapexone, mu = truth$lone),
  y2     = n_shared + rnbinom(n, size = truth$shapextwo, mu = truth$ltwo),
  vessel = factor(vessel)
)

c(cor = cor(dat$y1, dat$y2), var_y1 = var(dat$y1), var_y2 = var(dat$y2))
```

The pair is correlated because `n_shared` is drawn once and added to both
counts; drawing this component separately for each count would give two
independent counts with the same marginal distributions and no covariance. The
first count is the more variable of the two, as $\phi_{x1} < \phi_{x2}$
requires.

```{r}
#| label: disc-fit
#| eval: !expr has_stan
#| results: hide
#| message: false
fit <- brm(
  bf(y1 | vint(y2) ~ 1,
     mu ~ 1 + (1 | vessel),
     nlf(lambdaone ~ lamx), nlf(lambdatwo ~ lamx),
     lamx ~ 1, shapes ~ 1, shapexone ~ 1, shapextwo ~ 1, nl = TRUE),
  family   = binegbin(),
  stanvars = binegbin_stanvars(),
  data     = dat,
  prior    = c(
    prior(normal(2, 1),   class = "Intercept"),
    prior(normal(2, 1),   class = "b",         nlpar = "lamx"),
    prior(normal(0, 1.5), class = "Intercept", dpar  = "shapes"),
    prior(normal(0, 1.5), class = "Intercept", dpar  = "shapexone"),
    prior(normal(0, 1.5), class = "Intercept", dpar  = "shapextwo")
  ),
  chains  = 4, iter = 2000, warmup = 1000,
  seed    = 20260825, refresh = 0, init = 0.5,
  control = list(adapt_delta = 0.99, max_treedepth = 12),
  backend = backend
)
```

The prior on each log dispersion is `normal(0, 1.5)`, which spans roughly
$[0.05, 20]$ at two standard deviations and so contains all three true
values without favouring any of them.

```{r}
#| label: disc-recovery
#| eval: !expr has_stan
draws <- as.data.frame(fit)

qi <- function(x) c(median = median(x), lower = quantile(x, 0.025, names = FALSE),
                    upper = quantile(x, 0.975, names = FALSE))

recovery <- rbind(
  shapexone = qi(exp(draws$b_shapexone_Intercept)),
  shapextwo = qi(exp(draws$b_shapextwo_Intercept)),
  shapes    = qi(exp(draws$b_shapes_Intercept))
)
knitr::kable(
  cbind(truth = c(truth$shapexone, truth$shapextwo, truth$shapes), recovery),
  digits = 2
)
```

Each interval contains its true value. That much the vignette already
established for a symmetric design. The claim particular to this section is the
next one:

```{r}
#| label: disc-logratio
#| eval: !expr has_stan
lr    <- draws$b_shapexone_Intercept - draws$b_shapextwo_Intercept
lr_ci <- quantile(lr, c(0.025, 0.975), names = FALSE)

c(true_log_ratio = log(truth$shapexone) - log(truth$shapextwo),
  median         = median(lr),
  lower          = lr_ci[1],
  upper          = lr_ci[2])
```

The 95% posterior interval for $\log \phi_{x1} - \log \phi_{x2}$ excludes
zero, so the data separate the two dispersions rather than the priors: both
priors are identical, and centred at a log ratio of zero.

```{r}
#| label: disc-diagnostics
#| eval: !expr has_stan
c(divergences = sum(nuts_params(fit, pars = "divergent__")$Value),
  max_rhat    = max(rhat(fit), na.rm = TRUE))
```

```{r}
#| label: dispersion-posteriors
#| eval: !expr has_stan
#| fig.width: 8
#| fig.height: 3.4
#| fig.alt: >
#|   Two panels. The left shows posterior densities for the two excess
#|   dispersions on the log scale, well separated, with their true values
#|   marked. The right shows the posterior for their log ratio, with zero
#|   marked outside the bulk of the distribution.
op <- par(mfrow = c(1, 2), mar = c(4.1, 4.1, 2.6, 1.1), bty = "n")

d1 <- density(draws$b_shapexone_Intercept)
d2 <- density(draws$b_shapextwo_Intercept)
plot(d1, xlim = range(d1$x, d2$x), ylim = c(0, max(d1$y, d2$y)),
     lwd = 2, col = "#C4622D", main = "The two excess dispersions",
     xlab = expression(log ~ phi), ylab = "density")
lines(d2, lwd = 2, col = "#2D6A7F")
abline(v = log(truth$shapexone), lty = 3, col = "#C4622D")
abline(v = log(truth$shapextwo), lty = 3, col = "#2D6A7F")
legend("topright", bty = "n", lwd = 2, cex = 0.85,
       col = c("#C4622D", "#2D6A7F"),
       legend = c("shapexone", "shapextwo"))

plot(density(lr), lwd = 2, col = "#8A8072", main = "Their log ratio",
     xlab = expression(log ~ phi[x1] - log ~ phi[x2]), ylab = "density")
abline(v = 0, lwd = 2)
abline(v = log(truth$shapexone) - log(truth$shapextwo), lty = 3)
mtext("solid = zero, dotted = truth", side = 3, line = -1.1,
      cex = 0.7, adj = 0.97)

par(op)
```

Under partial observation, the same demonstration requires a majority of rows to
be matched, because `shapexone` is then informed by those rows alone.
`tests/testthat/test-binegbin-dispersions.R` runs both designs.

## Quantities returned by `posterior_epred()` and `posterior_predict()`

`posterior_epred()` returns

$$
\mathrm{E}[y_1 \mid y_2] = \mathrm{E}[N_{\text{shared}} \mid y_2] + \lambda_1,
$$

the expected first count given the second count observed on that row. The
returned quantity is a conditional expectation rather than a marginal one, and
the conditioning is on data rather than on a fitted value.

For `bipois()`, the conditional distribution is available in closed form:
conditioning a sum of independent Poisson counts on its total gives a Binomial,
so $N_{\text{shared}} \mid y_2 \sim \mathrm{Binomial}(y_2,\ \mu/(\mu +
\lambda_2))$ and the expectation is $y_2\,\mu/(\mu + \lambda_2) +
\lambda_1$. A sum of negative-binomial counts has no Binomial conditional, so
`binegbin()` evaluates the discrete conditional law over
$k = 0, \ldots, y_2$ and sums it. Both are exact, and each agrees with the
mean of that family's own `posterior_predict()` draws to within Monte Carlo
error.

`posterior_epred(fit)`, and therefore `fitted()` and `conditional_effects()`,
dispatch to the family's own method in the ordinary way. Neither family
supports `resp_trunc()`, so the brms limitation by which `posterior_epred()`
errors on a truncated custom family cannot arise here.

## Checks performed

The test suite is run with and without a Stan toolchain; the toolchain-free run
skips every model fit. For the likelihood, the suite verifies:

- the Stan implementation against an independent brute-force reference written
  in R, agreeing to about $10^{-14}$ across a grid of rates and dispersions
  and at edge cases;
- normalisation to 1, over the pair on the matched branch and over $y_2$ on
  the unmatched one;
- the moment identities above, by exact summation over the joint probability
  mass function rather than by simulation;
- the marginal identity, that summing the matched branch over $y_1$
  reproduces the unmatched branch;
- the Poisson limit, checked as a limit — the discrepancy must shrink as
  $\phi$ grows — rather than at a single tolerance.

For the post-processing methods, the suite verifies that each expectation equals the
mean of its own predictive draws, that the observation flag changes neither the
prediction nor the expectation, and that exchanging the two excess dispersions
anywhere they are routed produces a detectably different answer.

## References
