---
title: "A worked partially observed fit"
description: >
  Simulating a design in which the first count is recorded on only some rows,
  fitting it, and scoring the imputed first count against values withheld from
  the model.
bibliography: ../references.bib
link-citations: true
---

```{r}
#| label: setup
#| include: false
# This article is PRECOMPILED (see vignettes/precompile.R) because it fits a
# model. Figures are generated once, here, and committed to
# vignettes/articles/figure/ alongside the .Rmd.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  dev      = "svg",
  fig.path = "figure/"
)
has_stan <- requireNamespace("cmdstanr", quietly = TRUE) ||
  requireNamespace("rstan", quietly = TRUE)
backend <- if (requireNamespace("cmdstanr", quietly = TRUE)) "cmdstanr" else "rstan"
```

```{r}
#| label: libraries
#| message: false
#| warning: false
library(brms)
library(bicountbrms)
```

A partially paired design records the second count on every row and the first
count on only some of them. `binegbin_partialobs()` admits both kinds of row
under one likelihood: a matched row contributes the full joint probability of
the pair, and a row whose first count was never recorded contributes the second
count's marginal, taken from that same joint model.

Partial observation is not censoring in the sense of brms's `cens()` addition
term, which describes a value known to lie in a set. Here the first count was
not observed at all, and the likelihood marginalises over its whole support.

## Simulating a design with 40% of first counts unrecorded

The complete pair is generated first, and the first count is then withheld on
the unmatched rows. Keeping the withheld values aside makes it possible to score
the model's imputation against them at the end, which is not something a real
dataset permits.

```{r}
#| label: sim
set.seed(20260826)

n <- 400L
truth <- list(
  mu        = 6,
  lambdaone = 4,
  lambdatwo = 5,
  shapes    = 3,
  shapexone = 2,
  shapextwo = 4
)

n_shared <- rnbinom(n, size = truth$shapes,    mu = truth$mu)
y1_full  <- n_shared + rnbinom(n, size = truth$shapexone, mu = truth$lambdaone)
y2       <- n_shared + rnbinom(n, size = truth$shapextwo, mu = truth$lambdatwo)

# 60% of rows have both counts; on the rest the first count is withheld.
y1_obs <- rbinom(n, 1L, 0.6)

dat <- data.frame(
  y1     = ifelse(y1_obs == 1L, y1_full, 0L),
  y2     = y2,
  y1_obs = y1_obs
)

withheld <- y1_full[y1_obs == 0L]
c(n = n, matched = sum(y1_obs), unmatched = sum(1L - y1_obs))
```

Two details of the data matter.

The observation flag is a 0/1 integer column, supplied through `vint()`
alongside the second count: `vint(y2, y1_obs)` binds `vint1` to the second
count and `vint2` to the flag, in the order they are listed.

On an unmatched row, the response column is not read by the likelihood, so any
non-negative integer serves as a placeholder; `0` is used above. Do not write
`NA` there. brms drops rows with a missing response before the family is
reached, which would discard the second count on exactly the rows the family
exists to use.

## Fitting

```{r}
#| label: fit
#| eval: !expr has_stan
#| results: hide
#| message: false
fit <- brm(
  bf(y1 | vint(y2, y1_obs) ~ 1,
     mu ~ 1,
     lambdaone ~ 1, lambdatwo ~ 1,
     shapes ~ 1, shapexone ~ 1, shapextwo ~ 1),
  family   = binegbin_partialobs(),
  stanvars = binegbin_partialobs_stanvars(),
  data     = dat,
  prior    = c(
    prior(normal(2, 1),   class = "Intercept"),
    prior(normal(2, 1),   class = "Intercept", dpar = "lambdaone"),
    prior(normal(2, 1),   class = "Intercept", dpar = "lambdatwo"),
    prior(normal(0, 1.5), class = "Intercept", dpar = "shapes"),
    prior(normal(0, 1.5), class = "Intercept", dpar = "shapexone"),
    prior(normal(0, 1.5), class = "Intercept", dpar = "shapextwo")
  ),
  chains  = 4, iter = 2000, warmup = 1000,
  seed    = 20260826, refresh = 0,
  control = list(adapt_delta = 0.95),
  backend = backend
)
```

```{r}
#| label: diagnostics
#| eval: !expr has_stan
c(divergences = sum(nuts_params(fit, pars = "divergent__")$Value),
  max_rhat    = max(rhat(fit), na.rm = TRUE))
```

## The contribution of an unmatched row

An unmatched row is scored by

$$
P(y_2 = y) = \sum_{k=0}^{y}
  \mathrm{NB2}(k \mid \mu, \phi_{\text{s}})\,
  \mathrm{NB2}(y - k \mid \lambda_2, \phi_{x2}),
$$

which is the joint model with the first count summed out over its whole
support. The row is neither dropped nor given a separate univariate model on
the second count, either of which would be a different model from the one the
matched rows are fitted to.

Two consequences follow. An unmatched row informs $\mu$,
$\phi_{\text{s}}$, $\lambda_2$, $\phi_{x2}$ and any group-level
effects. It does not inform $\lambda_1$ or $\phi_{x1}$, which appear
only in the matched branch.

Which parameters an unmatched row informs is a property of the likelihood rather
than of this fit. What that property implies for a particular design is taken up
further down.

```{r}
#| label: recovery
#| eval: !expr has_stan
draws <- as.data.frame(fit)

pars <- c(mu        = "b_Intercept",
          lambdaone = "b_lambdaone_Intercept",
          lambdatwo = "b_lambdatwo_Intercept",
          shapes    = "b_shapes_Intercept",
          shapexone = "b_shapexone_Intercept",
          shapextwo = "b_shapextwo_Intercept")

recovery <- do.call(rbind, lapply(names(pars), function(p) {
  x <- exp(draws[[pars[[p]]]])
  data.frame(
    dpar     = p,
    informed = if (p %in% c("lambdaone", "shapexone")) "matched rows" else "every row",
    truth    = truth[[p]],
    median   = median(x),
    lower    = unname(quantile(x, 0.025)),
    upper    = unname(quantile(x, 0.975))
  )
}))

knitr::kable(recovery, digits = 2, row.names = FALSE)
```

Five of the six intervals contain their true value. The sixth,
$\phi_{x1}$, does not: its interval runs from 2.03 to 10.07 against a truth
of 2, so it misses at the lower end by 0.03.

The seed was not changed to make the miss disappear. A 95% interval from a
correctly specified model excludes the truth 5% of the time by construction,
and across six parameters the chance that at least one interval misses is about
26%. One near miss in six is therefore the expected behaviour of a working
implementation, not evidence against it. The calibration evidence in this
article is the coverage check further down, which is taken over
`r length(withheld)` withheld counts rather than over one fit.

The three dispersions have much the wider intervals, because each is estimated
from an aggregate mean–variance mismatch rather than from any directly observed
quantity.

## Imputing the withheld count

`posterior_predict()` returns a draw of the first count for every row,
conditional on the second count recorded there. The observation flag selects a
branch of the likelihood; it does not change the prediction. Rows on which the
first count was never recorded are therefore predicted in exactly the same way
as matched rows, which is what the imputation depends on.

```{r}
#| label: impute
#| eval: !expr has_stan
pp  <- posterior_predict(fit)
idx <- which(dat$y1_obs == 0L)

imp <- data.frame(
  withheld = withheld,
  median   = apply(pp[, idx, drop = FALSE], 2, median),
  lower    = apply(pp[, idx, drop = FALSE], 2, quantile, 0.05),
  upper    = apply(pp[, idx, drop = FALSE], 2, quantile, 0.95)
)
imp$covered <- imp$withheld >= imp$lower & imp$withheld <= imp$upper

c(n_unmatched      = nrow(imp),
  empirical_cover  = mean(imp$covered),
  nominal          = 0.90)
```

The empirical coverage of the nominal 90% interval is a check on the
imputation, not a test of the method: simulating from the model and refitting
it makes the coverage nominal by construction when the implementation is
correct. It is reported because a departure would indicate an implementation
error, and because the interval widths themselves are the quantity a user
planning such a design wants to see.

```{r}
#| label: imputation-check
#| eval: !expr has_stan
#| fig.width: 8
#| fig.height: 4.2
#| fig.alt: >
#|   Withheld first counts on the horizontal axis against their predicted
#|   values on the vertical, with 90% predictive intervals as vertical
#|   segments and a one-to-one line. The points follow a flatter trend than
#|   the line, lying above it at low counts and below it at high counts.
#|   Most intervals cross the line.
op <- par(mar = c(4.1, 4.1, 2.6, 1.1), bty = "n")
ord <- order(imp$withheld)
plot(imp$withheld[ord], imp$median[ord],
     xlim = range(c(imp$withheld, imp$lower, imp$upper)),
     ylim = range(c(imp$withheld, imp$lower, imp$upper)),
     pch = 16, cex = 0.7, col = "#2D6A7F",
     xlab = "withheld first count", ylab = "predicted first count",
     main = "Imputation against the values withheld from the model")
segments(imp$withheld[ord], imp$lower[ord], imp$withheld[ord], imp$upper[ord],
         col = "#2D6A7F44")
abline(0, 1, lwd = 2, col = "#C4622D")
mtext("segments = 90% predictive interval; line = one to one",
      side = 3, line = -1.1, cex = 0.75, adj = 0.97)
par(op)
```

```{r}
#| label: shrinkage-numbers
#| include: false
# Quoted in the prose below; computed so the text tracks the fit. `imp` exists
# only where the fitting chunks ran, so a no-Stan preview gets NA rather than
# an error.
marg_mean <- truth$mu + truth$lambdaone
imp_slope <- if (exists("imp")) {
  unname(coef(stats::lm(median ~ withheld, data = imp))[2])
} else NA_real_
```

The prediction is conditional on the second count, not on the withheld first
count. The two counts have only the shared component in common, so the second
count constrains the first only through that component and the prediction is
shrunk towards the marginal mean of the first count, $\mu + \lambda_1$,
here `r marg_mean`. High withheld counts are therefore under-predicted and low
ones over-predicted: the posterior median regressed on the withheld count has
slope `r round(imp_slope, 2)` rather than 1. The vertical spread measures what
the second count leaves undetermined about the first.

## Choosing a matched fraction

$\lambda_1$ and $\phi_{x1}$ appear only in the matched branch, so as
the matched count approaches zero they stop being identified and their
posteriors return their priors. $\lambda_2$ continues to be estimated from
the second counts however few pairs are matched.

The matched rows also do work for the other four parameters, which the branch
structure alone does not show. An unmatched row constrains $\mu$ and
$\lambda_2$ only through the convolution that gives the second count's
marginal, and that convolution determines the sum of the two components better
than it determines the division between them. Separating the shared component
from the second source's own is done by the matched rows, because only a
matched row observes both counts of a pair. For `bipois_partialobs()`, this is
sharper still: an unmatched row involves $\mu$ and $\lambda_2$ through
their sum $\mu + \lambda_2$ exactly, with no convolution to separate them.

Where the between-source bias or the congruence is the quantity of interest,
the matched subset is the effective sample size for it. That is worth knowing
before the data are collected.

## Telling which constructor a stored fit used

`family$name` is `"binegbin"` for both constructors, so it does not
distinguish them. The presence of the second supplementary integer does:

```{r}
#| label: which-shape
#| eval: !expr has_stan
c(has_vint2 = "vint2" %in% names(standata(fit)),
  vars      = paste(fit$family$vars, collapse = ", "))
```

## References
