---
title: "Migration and errata"
description: >
  What each release changed, what to substitute in existing code, what to do
  with a fit stored under a name that no longer exists, and one numerical
  result that was wrong before 0.9.0.
bibliography: ../references.bib
link-citations: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Package history

These families were first released as `skellambrms` (0.1.0–0.5.0) and then as
`pairedcountbrms` (0.6.0–0.8.0). That second package held two unrelated suites:
families modelling a count pair jointly, and families modelling its difference
$d = y_1 - y_2$ directly. The two shared no code. Release 0.9.0 separated
them. The joint families became this package; the difference families returned
to [`skellambrms`](https://github.com/anhsmith/skellambrms).

The separation renamed nothing, so `library(pairedcountbrms)` becomes
`library(bicountbrms)` or `library(skellambrms)` according to which suite was in
use. `github.com/anhsmith/pairedcountbrms` redirects here.

## Changes by release

**0.8.0** replaced the single excess dispersion `shapex` with the per-margin
pair `shapexone`/`shapextwo`, for the partially observed family alone.

**0.9.0** renamed `binegbin_joint()` to `binegbin_cens()`, on the grounds that
every family in the package modelled the pair jointly and `_joint` therefore
distinguished nothing. The same release added `bipois_cens()`.

**0.10.0** replaces four families with two. Each of `bipois` and `binegbin` now
has two constructors: a plain one for a fully paired design, and a
`_partialobs()` one for a design in which the first count is missing on some
rows. Both constructors of a pair return the same `custom_family` name, so one
Stan function and one set of post-processing methods serve both. The same
release gives `binegbin()` the six distributional parameters previously held by
`binegbin_cens()` alone.

## Substitutions for existing code

| up to 0.9.1 | from 0.10.0 |
|---|---|
| `bipois_cens()` | `bipois_partialobs()` |
| `binegbin_cens()`, `binegbin_joint()` | `binegbin_partialobs()` |
| `bipois_cens_stanvars()` | `bipois_partialobs_stanvars()` |
| `binegbin_cens_stanvars()`, `binegbin_joint_stanvars()` | `binegbin_partialobs_stanvars()` |
| `binegbin()` with `shapex ~ 1` | `binegbin()` with `shapexone ~ 1, shapextwo ~ 1` |

The old names are removed outright. There is no deprecation layer, and nothing
is left to remove at a later release.

A formula written against the five-dpar `binegbin()` needs its dispersion term
replaced. Either estimate the two separately:

```{r}
#| label: untied
bf(y1 | vint(y2) ~ 1, mu ~ 1,
   nlf(lambdaone ~ lamx), nlf(lambdatwo ~ lamx),
   lamx ~ 1, shapes ~ 1, shapexone ~ 1, shapextwo ~ 1, nl = TRUE)
```

or reproduce the pre-0.10.0 model exactly, by routing both through one
non-linear parameter:

```{r}
#| label: tied
bf(y1 | vint(y2) ~ 1, mu ~ 1,
   nlf(lambdaone ~ lamx), nlf(lambdatwo ~ lamx),
   nlf(shapexone ~ shapexx), nlf(shapextwo ~ shapexx),
   lamx ~ 1, shapes ~ 1, shapexx ~ 1, nl = TRUE)
```

The second form is term for term the likelihood the single `shapex` gave, and a
package test verifies that the generated Stan assigns both distributional
parameters from one non-linear parameter.

A prior moves with it. Under the tied form, the dispersion prior is written

```{r}
#| label: tied-prior
prior(normal(0, 1.5), class = "b", nlpar = "shapexx")
```

rather than the pre-0.10.0 `class = "Intercept", dpar = "shapex"`. Both fields
change. A prior written the old way names no parameter in the new model and is
dropped without a warning, which leaves the dispersion improper. See
[Choosing priors](choosing-priors.html).

## `_cens` named the wrong mechanism

brms already uses `cens()` as an addition term meaning a value known to lie in
a set — `left`, `right`, `interval` — and supplies no code for a value that was
not observed at all. On the rows these
constructors admit, the first count was not recorded, and the likelihood
marginalises over its whole support rather than over a bounded set. A brms user
meeting `_cens` had every reason to map one mechanism onto the other.

`partialobs` names what is partial, which is the *pair*, and which is a property
of the data rather than of the model. `unobs` and `unmatched` name only the
minority class of rows; `partial` alone collides with partial likelihood and
partial pooling.

## Fits stored under a removed name

A `brmsfit` stores its own family object, and brms builds the name of each
post-processing method from the stored family name at the time of the call. A
fit made with `bipois_cens()`, `binegbin_cens()` or `binegbin_joint()` therefore
looks for `log_lik_bipois_cens()` and its counterparts, which 0.10.0 does not
define. No argument to `loo()` or `posterior_predict()` changes which name is
looked up.

To post-process such a fit, install the last release that defined those methods
and keep it attached for that work:

```{r}
#| label: pin
pak::pak("anhsmith/bicountbrms@v0.9.1")
```

0.10.0 ships no forwarding functions, deliberately. Pinning the whole package
is a stronger guarantee than a hand-picked set of forwarders, which would then
have to be kept correct as the families continue to change.

## Stored `binegbin` fits require no action

A fit stored under the name `binegbin` is the exception. It declares the name
0.10.0 keeps, so it is post-processed by the current methods whatever version is
pinned, and it works. Such fits exist in quantity: a scan of the project this
package was written for found 103 of them, each declaring the pre-0.7.0 rate
names `lambdaem` and `lambdalb` together with the single `shapex`.

Three fallbacks apply together to make that work.

1. `.get_rate()` resolves `lambdaem`/`lambdalb` to the `lambdaone`/`lambdatwo`
   positions, so a fit predating the 0.7.0 rename reads correctly.
2. `.SHAPEXONE_NAMES` and `.SHAPEXTWO_NAMES` both list `shapex` last, so a fit
   with one excess dispersion resolves it to both per-margin positions. That is
   precisely the constraint $\phi_{x1} = \phi_{x2}$ the five-dpar family
   imposed.
3. An absent `vint2` selects the matched branch of the likelihood, which is
   correct for a fit made with a constructor that had no observation flag.

`tests/testthat/test-dpar-compat.R` builds that exact combination and checks all
three post-processing methods against the six-dpar answer with the two
dispersions tied.

## Telling which constructor a stored fit used

From 0.10.0 `family$name` is identical for both constructors of a family, so it
does not distinguish them. The presence of the second supplementary integer
does:

```{r}
#| label: which-shape
"vint2" %in% names(standata(fit))   # TRUE for a partially observed fit
fit$family$vars                     # c("vint1[n]", "vint2[n]") or c("vint1[n]", "1")
```

## Erratum: `posterior_epred_binegbin()` before 0.9.0

Releases before 0.9.0 computed `posterior_epred_binegbin()` using the
*marginal* shared fraction $\mu/(\mu + \lambda_2)$ in place of the
*conditional* one. That substitution is `bipois`'s answer, exact only in the
Poisson limit, and it made the expectation disagree with the family's own
`posterior_predict()` draws by more than Monte Carlo error.

Over a grid of plausible rates and dispersions, with $y_2$ within one
standard deviation of its mean, the substituted value was in error by more than
5% in about half the settings and by more than 20% in a third of them. The
direction of the error depends on which component is the more dispersed.

`log_lik()`, `loo()` and `posterior_predict()` were not affected at any release.
Any number computed with an earlier `posterior_epred_binegbin()` should be
recomputed.

## Erratum: the dispersion routing in `posterior_predict()`, before 0.9.1

Release 0.9.1 added `tests/testthat/test-cens-predict.R`, since superseded by
`test-partialobs-predict.R`, after a check of the test suite's discrimination.
In `posterior_predict_binegbin_cens()`, the conditional split of
$N_{\text{shared}} \mid y_2$ is weighted by `shapextwo`, while the fresh
source-specific count added on top takes `shapexone`. Exchanging those two
passed 412 assertions across six test files without a single failure, because
every prediction test then in the suite built its fixture from a single
five-dpar `shapex`, which both name vectors resolve to identically.

No released version computed the exchanged quantity. The defect was in the test
suite's power to detect an error, not in the shipped code.

## References
